---
output: github_document
bibliography: inst/REFERENCES.bib
link-citations: true
---

<!-- README.md is generated from README.Rmd. Edit the .Rmd and run devtools::build_readme(). -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  dev = "svg",
  out.width = "100%"
)
```

```{r helpers, include = FALSE}
tri_xy <- function(theta) {
  rbind(c(0.5, 0, 1), c(sqrt(3) / 2, 0, 0)) %*% as.matrix(theta)
}

plot_null <- function(
  null,
  points = NULL,
  weights = NULL,
  main = NULL,
  fill = "#c96f5b",
  bg = c("#2c6f9b", "white"),
  alpha = 0.55
) {
  corners <- tri_xy(diag(3))
  op <- par(mar = c(0.5, 0.5, if (is.null(main)) 0.5 else 2, 0.5))
  on.exit(par(op), add = TRUE)
  plot.new()
  plot.window(c(-0.05, 1.05), c(-0.05, 0.92), asp = 1)
  for (s in parts(null@region)) {
    xy <- tri_xy(s@vertices)
    polygon(
      xy[1, ],
      xy[2, ],
      col = adjustcolor(fill, 0.45),
      border = fill,
      lwd = 1.5
    )
  }
  polygon(corners[1, ], corners[2, ], border = "grey50")
  text(
    corners[1, ],
    corners[2, ],
    c(1, 2, 3),
    pos = c(3, 1, 1),
    cex = 0.8,
    col = "grey40"
  )
  if (!is.null(points)) {
    groups <- if (is.list(points)) points else list(points)
    sizes <- if (is.list(weights)) weights else list(weights)
    # Marker area proportional to mixing weight -- `cex` scales a linear
    # dimension, so the honest map is sqrt -- on a scale shared across groups,
    # so a heavy atom looks heavy wherever it sits.
    all_w <- unlist(sizes)
    for (g in seq_along(groups)) {
      xy <- tri_xy(groups[[g]])
      w <- sizes[[min(g, length(sizes))]]
      cex <- if (is.null(w)) 1.2 else pmax(0.35, 1.7 * sqrt(w / max(all_w)))
      points(
        xy[1, ],
        xy[2, ],
        pch = 21,
        bg = adjustcolor(bg[[(g - 1L) %% length(bg) + 1L]], alpha),
        col = "grey15",
        cex = cex,
        lwd = 1.1
      )
    }
  }
  if (!is.null(main)) {
    title(main, font.main = 1, cex.main = 1)
  }
  invisible(NULL)
}
```

# ripr

`ripr` computes reverse information projections onto nulls that take the form of a union of convex parts, and then proves an upper bound on the expectation over the null of whatever random variable you get.
The projection is fitted via Frank--Wolfe [@Jaggi2013] and/or EM; bounds are proved by a branch-and-bound algorithm (currently for multinomial families, via the Bernstein enclosure of @Garloff1986 and @Leroy2012).
At the end, you get a random variable that is genuinely an e-variable for the null.

Fitting and certification are deliberately separate: **any** approximate projection can be rescaled (by an upper bound on the expectation over the null) to get a valid e-variable. This way, a very difficult, non-convex optimisation problem splits into two easier problems: approximate a distribution, then compute a bound. The better the fit, the higher the e-power, and validity is guaranteed.

## Installation

```{r install, eval = FALSE}
# install.packages("remotes")
remotes::install_github("fleverest/ripr")
```

## The usual workflow

Take a three-category multinomial distribution (`n = 20` samples) and the plurality null: candidate 1 does not win outright, `H0 = union_j {theta : theta_1 <= theta_j}`.
This null is a union of two convex parts, so it's built with `union()`:

```{r problem, message = FALSE}
library(ripr)

K <- 3L
n <- 20L
family <- multinomial_family(n_trials = n, k = K)

# Defines {theta : theta_1 <= theta_j}
plurality_part <- function(j) {
  vertices <- diag(K)
  vertices[, 1L] <- replace(numeric(K), c(1L, j), 0.5)
  simplex_region(vertices = vertices)
}

plurality <- null_model(
  family,
  union(plurality_part(2), plurality_part(3))
)
```

We nominate a discrete mixture-multinomial alternative (here a single dirac at `q = (0.40, 0.35, 0.25)`), so `Q` is just `Multinomial(20, q)`:

```{r alternative}
Q <- family(c(0.40, 0.35, 0.25))
```

Fitting is a sequence of step verbs: `fw_step()` for Frank--Wolfe, `em_step()` for EM. Here we run Frank--Wolfe for 40 iterations, then `ripr_finish()` solves the weights and drops any low-mass atoms:

```{r fit}
set.seed(1)
state <- ripr_init(Q, plurality)
state <- fw_step(
  state,
  times = 40L,
  record_gap = TRUE,
  until = gap_below(1e-10)
)
fit <- ripr_finish(state, reoptimise = TRUE, identify = TRUE, record_gap = TRUE)

c(kl = fit$kl, gap = fit$gap_final, atoms = n_atoms(fit$W0))
```

```{r plot-plurality-fit}
#| echo: false
#| fig.width: 4.5
#| fig.height: 4
#| out.width: "60%"
#| fig.align: "center"
#| fig.cap: Plurality null, with the support of the alternative mixing distribution W1 and the fitted null mixture W0.
plot_null(
  plurality,
  points = list(atoms(fit$W0), atoms(Q@mixing)),
  weights = list(weights(fit$W0), weights(Q@mixing) / 4)
)
```

The fit gives a candidate e-variable: the mixture likelihood ratio `Q / P*`.
`random_variable`s are callable, mapping outcomes to their realisation:

```{r ratio}
X <- likelihood(Q, label = "Q") /
  likelihood(fit$P_star, label = "P*")

outcomes <- rbind(
  c(10L, 10L, 0L),
  c(8L, 7L, 5L)
)
X(outcomes)
```

`certify()` proves an upper bound on expectation of `X` over the null, including a lower-bound that is the largest value the search actually attained:

```{r certify}
cert <- certify(X, plurality, tol = 1e-9)
c(
  upper = cert$sup_ub,
  attained = cert$sup_lb,
  width = cert$sup_ub - cert$sup_lb
)
```

Rescaling by the upper bound turns `X` into a bona fide e-variable for the `plurality` null:

```{r evariable}
E <- X / cert$sup_ub
print(E)
print(E(outcomes))
```

## Learn more

`vignette("ripr")` walks through the same example in more depth: why validity survives a deliberately bad fit, how a bound from `certify()` differs from `sup_lb()`, and what happens when no certification method exists for a family/geometry pair.
`vignette("regions")` covers the region interface and set algebra (`union()`, `intersect()`, `setdiff()`, `disjoin()`, ...) used to build nulls out of convex pieces.

## References
