---
output: github_document
---

<!-- README.md is generated from README.Rmd. Edit the .Rmd and run devtools::build_readme(). -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

# ripr

`ripr` computes the **reverse information projection** (RIPr) `P*` of an
alternative distribution `Q` onto a (possibly non-convex, union-structured)
null hypothesis, together with the **duality-gap certificate** for the rescaled
e-variable `(Q / P*) / (1 + gap)`, whose expectation under every null
distribution is at most 1.

The projection is found by Frank--Wolfe and EM over a mixture of atoms on the
null faces. It requires the specification of a sampling `family` (the model
`p_theta`), the null geometry (a union of `face`s), and a surrogate `alternative`
`Q`. The alternative is an `outcome_distribution` — a law over the sample space —
which you can supply directly or build from a `mixing` measure over the
parameter space (e.g. `point_mixing`, `finite_mixing`), marginalised through the
family. These components are templated as
[S7](https://rconsortium.github.io/S7/) base classes, so new sample spaces drop
in without touching the algorithmic core. The algorithms require the ability to
compute KL divergences and gradients, which may be implemented either via
**exact** numerical methods or by **Monte Carlo** sampling. These two routes are
encapsulated in the `engine` class.


```{r install, eval = FALSE}
# install from github
renv::install("fleverest/ripr")
```


## Example 1: a one-sided binomial null

Let `X ~ Binomial(10, p)`, written as a 2-category multinomial with
`theta = (p, 1 - p)`. Consider testing `H_0: p <= 1/2` against `H_1: p > 1/2`,
adopting the simple alternative `Q = Binomial(10, 3/4)`. The RIPr of `Q` onto
the convex null is a **point mass at `p = 1/2`**. We demonstrate this using both
the exact and Monte Carlo engines.

```{r ex1, message = FALSE}
library(ripr)
n <- 10
fam <- multinomial_family(n_trials = n, k = 2)
# `ripr_problem` takes Q as an `outcome_distribution` over the sample space.
# Build one by marginalising a `mixing` measure over the parameter theta (here
# it is just a point mass at theta = (3/4, 1/4)) through the family, giving the
# outcome law Q = Binomial(10, 3/4).
Q <- as_marginal(point_mixing(theta_star = c(0.75, 0.25)), fam)

# H_0 = {p <= 1/2} is the simplex segment from p = 0 to the tie p = 1/2.
face <- polytope_face(vertices = cbind(c(0, 1), c(0.5, 0.5)), face_index = 1)
null <- null_region(faces = list(face))
```

**Exact engine** (enumerates all 11 outcomes):

```{r ex1-exact}
set.seed(1)
prob <- ripr_problem(fam, null, Q)
res <- run_ripr(
  prob,
  # Deliberately initialising somewhere sub-optimal for illustrative purposes
  init_atoms = matrix(c(0.25, 0.75), ncol = 1),
  init_atom_faces = 1L,
  fw_iters = 8,
  em_iters = 5,
  prune_threshold = 1e-6, # drop the numerically-dead atoms for output
  verbose = FALSE
)

cbind(
  p = res$projection@mixing@components[1, ],
  weight = res$projection@mixing@weights
)
```

`res$projection` is a `marginal` — the fitted `P*` as a distribution over
outcomes — and its `@mixing` is the `finite_mixing` holding the atoms and
weights on the null. `run_ripr()` returns the fitted e-variable directly; call
`e_value()` on new
data. Observing `(8, 2)` yields evidence against `H_0`, while a tie `(5, 5)`
yields evidence in favour of it:

```{r ex1-evalue}
res$e_variable
e_value(res$e_variable, rbind(c(8, 2), c(5, 5)))
```

**Monte Carlo engine** (sample 4000 draws from `Q`) reaches the same projection:

```{r ex1-mc}
eng_mc <- mc_engine(fam, Q, n_draws = 1000)
res_mc <- run_ripr(
  ripr_problem(fam, null, Q, engine = eng_mc),
  init_atoms = matrix(c(0.25, 0.75), ncol = 1),
  init_atom_faces = 1L,
  fw_iters = 10,
  em_iters = 10,
  gap_tol = 1e-3,
  verbose = FALSE
)

cbind(
  p = res_mc$projection@mixing@components[1, ],
  weight = res_mc$projection@mixing@weights
)
```

## Example 2: a 2-D Gaussian cone null (union of half-spaces)

Now let `X ~ N(mu, I_2)` and test the **cone alternative** `H_1: mu_1 > |mu_2|`
— equivalently `mu_1 > mu_2` **and** `mu_1 > -mu_2` — with the point alternative
`mu = (2, 1)`. Its complement, the null
`H_0 = {mu_1 <= mu_2} ∪ {mu_1 <= -mu_2}`, is a union of two half-spaces.

```{r ex2}
fam <- gaussian_family(dim = 2) # standard: sigma = I
mu <- c(2, 1)
Q <- as_marginal(point_mixing(theta_star = mu), fam)

faces <- list(
  halfspace_face(v = c(1, -1), c = 0, face_index = 1), # mu_1 <= mu_2
  halfspace_face(v = c(1, 1), c = 0, face_index = 2) # mu_1 <= -mu_2
)
null <- null_region(faces = faces)
```

The sample space is continuous here, so we use the Monte Carlo engine: 1000
draws to fit the mixture, and fresh `certify_draws = 5000` samples, independent
of the fit, to certify the result (drawn twice by default — see below).

```{r ex2-mc}
set.seed(2)
eng <- mc_engine(fam, Q, n_draws = 1000)
prob <- ripr_problem(fam, null, Q, engine = eng)
init <- cbind(
  init_point(faces[[1]], mu),
  init_point(faces[[2]], mu)
)

res <- run_ripr(
  prob,
  init_atoms = init,
  init_atom_faces = c(1, 2),
  fw_iters = 25,
  em_iters = 5,
  n_seeds = 50,
  gap_tol = 1e-3,
  certify_draws = 5000,
  verbose = FALSE
)

M <- res$projection@mixing@components
cbind(
  mu1 = M[1, ],
  mu2 = M[2, ],
  weight = round(res$projection@mixing@weights, 3)
)
```

The projection concentrates on the `mu_1 = mu_2` tie-point — the null boundary
nearest the alternative `mu = (2, 1)` — with a little mass on the `mu_1 = -mu_2`
face. The certificate reports the guaranteed e-value growth rate and, for a
Monte Carlo engine, the standard error of the estimated gap:

```{r ex2-cert}
c(
  gap = res$gap,
  gap_se = res$certificate$gap_se,
  growth_rate = res$certificate$growth_rate
)
```

### Re-certifying on a larger sample

The Monte Carlo certificate above is a stochastic estimate: `certify()` resamples
fresh draws from `Q` (it does *not* reuse the fit draws) and, by default,
split-samples — one sample to locate the worst-case direction `theta*`, a second
to estimate the gap at it, so the estimate carries no selection bias. Its
standard error shrinks like `1 / sqrt(n)`, so re-certifying the *same* fitted
projection on more draws tightens it, with no re-fitting — just raise `n_draws`:

```{r ex2-recert}
set.seed(9)
cert <- certify(res$projection, prob, n_draws = 50000)
c(new_gap = cert$gap, new_se = cert$gap_se)
```

`certify()` takes the fitted `projection` (or any `finite_mixing` of atoms on the
null) and sweeps the oracle over the null faces. It resamples the problem's own
engine by default; pass `engine =` to certify against a different one (say an
`exact_engine`), and `split = FALSE` for a single-sample estimate.
