---
output: github_document
---

<!-- README.md is generated from README.Rmd. Edit the .Rmd and run devtools::build_readme(). -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

# ripr

`ripr` computes the **reverse information projection** (RIPr) `P*` of an
alternative distribution `Q` onto a (possibly non-convex, union-structured)
null hypothesis, together with the **duality-gap certificate** for the rescaled
e-variable `(Q / P*) / (1 + gap)`. With an exact engine the rescaled e-variable's
expectation under every null distribution is at most 1. With a Monte Carlo engine
the gap is estimated, so the bound is stochastic: it holds at level `conf`
(default 0.95), at the worst-case `theta*` the certifying sample selects.

The projection is found by Frank--Wolfe and EM over a mixture of atoms on the
null faces. It requires the specification of a sampling `family` (the model
`p_theta`), the null geometry (a union of `face`s), and a surrogate `alternative`
`Q`. The alternative is an `outcome_distribution` — a law over the sample space —
which you can supply directly or build from a `mixing` measure over the
parameter space (e.g. `point_mixing`, `finite_mixing`), marginalised through the
family. These components are templated as
[S7](https://rconsortium.github.io/S7/) base classes, so new sample spaces drop
in without touching the algorithmic core. The algorithms require the ability to
compute KL divergences and gradients, which may be implemented either via
**exact** numerical methods or by **Monte Carlo** sampling. These two routes are
encapsulated in the `engine` class.


```{r install, eval = FALSE}
# install from github
renv::install("fleverest/ripr")
```


## Example 1: a one-sided binomial null

Let `X ~ Binomial(10, p)`, written as a 2-category multinomial with
`theta = (p, 1 - p)`. Consider testing `H_0: p <= 1/2` against `H_1: p > 1/2`,
adopting the simple alternative `Q = Binomial(10, 3/4)`. The RIPr of `Q` onto
the convex null is a **point mass at `p = 1/2`**. We demonstrate this using both
the exact and Monte Carlo engines.

```{r ex1, message = FALSE}
devtools::load_all()
n <- 10
fam <- multinomial_family(n_trials = n, k = 2)
# `ripr_problem` takes Q as an `outcome_distribution` over the sample space.
# Build one by marginalising a `mixing` measure over the parameter theta (here
# it is just a point mass at theta = (3/4, 1/4)) through the family, giving the
# outcome law Q = Binomial(10, 3/4).
Q <- as_marginal(point_mixing(theta_star = c(0.75, 0.25)), fam)

# H_0 = {p <= 1/2} is the simplex segment from p = 0 to the tie p = 1/2.
face <- polytope_face(vertices = cbind(c(0, 1), c(0.5, 0.5)), face_index = 1)
null <- null_region(faces = list(face))
```

**Exact engine** (enumerates all 11 outcomes):

```{r ex1-exact}
set.seed(1)
prob <- ripr_problem(fam, null, Q)
res <- run_ripr(
  prob,
  # Deliberately initialising somewhere sub-optimal for illustrative purposes
  init_atoms = matrix(c(0.25, 0.75), ncol = 1),
  init_atom_faces = 1L,
  fw_iters = 8,
  em_iters = 5,
  prune_threshold = 1e-6, # drop the numerically-dead atoms for output
  verbose = FALSE
)

cbind(
  p = res$projection@mixing@components[1, ],
  weight = res$projection@mixing@weights
)
```

`res$projection` is a `marginal` — the fitted `P*` as a distribution over
outcomes — and its `@mixing` is the `finite_mixing` holding the atoms and
weights on the null. `run_ripr()` returns the fitted e-variable directly; call
`e_value()` on new
data. Observing `(8, 2)` yields evidence against `H_0`, while a tie `(5, 5)`
yields evidence in favour of it:

```{r ex1-evalue}
res$e_variable
e_value(res$e_variable, rbind(c(8, 2), c(5, 5)))
```

**Monte Carlo engine** (sample 4000 draws from `Q`) reaches the same projection:

```{r ex1-mc}
eng_mc <- mc_engine(fam, Q, n_draws = 1000)
res_mc <- run_ripr(
  ripr_problem(fam, null, Q, engine = eng_mc),
  init_atoms = matrix(c(0.25, 0.75), ncol = 1),
  init_atom_faces = 1L,
  fw_iters = 10,
  em_iters = 10,
  gap_tol = 1e-3,
  verbose = FALSE
)

cbind(
  p = res_mc$projection@mixing@components[1, ],
  weight = res_mc$projection@mixing@weights
)
```

## Example 2: a 2-D Gaussian cone null (union of half-spaces)

Now let `X ~ N(mu, I_2)` and test the **cone alternative** `H_1: mu_1 > |mu_2|`
— equivalently `mu_1 > mu_2` **and** `mu_1 > -mu_2` — with the point alternative
`mu = (2, 1)`. Its complement, the null
`H_0 = {mu_1 <= mu_2} ∪ {mu_1 <= -mu_2}`, is a union of two half-spaces.

```{r ex2}
fam <- gaussian_family(dim = 2) # standard: sigma = I
mu <- c(2, 1)
Q <- as_marginal(point_mixing(theta_star = mu), fam)

faces <- list(
  halfspace_face(v = c(1, -1), c = 0, face_index = 1), # mu_1 <= mu_2
  halfspace_face(v = c(1, 1), c = 0, face_index = 2) # mu_1 <= -mu_2
)
null <- null_region(faces = faces)
```

The sample space is continuous here, so we use the Monte Carlo engine: 1000
draws to fit the mixture, and fresh `certify_draws` samples, independent of the
fit, to certify the result. Certification is a single oracle sweep rather than a
per-iteration cost, so it is cheap to certify on far more draws than were used
to fit.

```{r ex2-mc}
set.seed(2)
eng <- mc_engine(fam, Q, n_draws = 1000)
prob <- ripr_problem(fam, null, Q, engine = eng)
init <- cbind(
  init_point(faces[[1]], mu),
  init_point(faces[[2]], mu)
)

res <- run_ripr(
  prob,
  init_atoms = init,
  init_atom_faces = c(1, 2),
  fw_iters = 25,
  em_iters = 5,
  n_seeds = 50,
  gap_tol = 1e-3,
  certify_draws = 1e5,
  verbose = FALSE
)

M <- res$projection@mixing@components
cbind(
  mu1 = M[1, ],
  mu2 = M[2, ],
  weight = round(res$projection@mixing@weights, 3)
)
```

The projection concentrates on the `mu_1 = mu_2` tie-point — the null boundary
nearest the alternative `mu = (2, 1)` — with a little mass on the `mu_1 = -mu_2`
face.

`run_ripr()` returns six elements, partitioned by where each number came from:
`projection` and `e_variable` are the deliverables; `certificate` holds
everything measured on the fresh certification sample; `history` and
`checkpoints` hold everything measured on the fit sample; `converged` reports
whether the *fit* gap met `gap_tol`.

```{r ex2-str}
names(res)
```

The certificate reports the guaranteed e-value growth rate, the standard error
of the estimated gap, and the settings that produced it:

```{r ex2-cert}
c(
  gap = res$certificate$gap,
  gap_se = res$certificate$gap_se,
  gap_used = res$certificate$gap_used,
  growth_rate = res$certificate$growth_rate,
  n_draws = res$certificate$n_draws
)
```

The fit record is one row per outer iteration, with the inner init/FW/EM steps
and the oracle argmax nested as list columns:

```{r ex2-history}
head(res$history[, c("iter", "gap", "gap_se", "support_size", "kl_after_em")], 4)
res$history$kl_trace[[2]] # the FW + EM steps inside iteration 1
```

`checkpoints$final` is always present and describes the returned `projection`
after pruning. Its three fields are exactly the arguments needed to resume the
fit, so a run can be continued without re-deriving anything:

```{r ex2-final}
fin <- res$checkpoints$final
str(fin, max.level = 1)

resume_kl <- function(w) {
  run_ripr(
    prob,
    init_atoms = do.call(cbind, fin$atoms),
    init_atom_faces = fin$atom_face_idx,
    init_weights = w,
    fw_iters = 0, em_iters = 0, # measure the starting point only
    certify_ess_min = 0, # not certifying here; see the accuracy note below
    verbose = FALSE
  )$history$kl_trace[[1]]$kl[1]
}
c(with_weights = resume_kl(fin$weights), uniform = resume_kl(NULL))
```

`init_weights` is what makes the resume a true continuation: without it the
atoms carry over but the mixture restarts from uniform weights, and EM has to
rediscover them.

This problem has a known closed form, so the cost of the certificate is
measurable. The exact RIPr is two atoms, at `(1.548311, 1.548311)` with weight
`0.95099` and `(1.267288, -1.267288)` with weight `0.04901`, giving
`KL* = 0.231375`. The shortfall of `growth_rate` below `KL*` is what the
certificate gives up: part optimisation slack, part Monte Carlo inflation.

### Re-certifying on a larger sample

The Monte Carlo certificate is a stochastic bound, and deliberately a
conservative one. `certify()` resamples fresh draws from `Q` (it does *not*
reuse the fit draws) and takes the oracle maximum over the null faces; that
maximum is biased *upward* for the true `sup_theta G`, which is the safe
direction, and `gap_used` inflates it further to a one-sided bound at level
`conf`. **`gap_used` is the only quantity that may be used to rescale an
e-variable.** The inflation term shrinks like `1 / sqrt(n)`, so re-certifying the
*same* fitted projection on more draws sharpens the certificate, with no
re-fitting — just raise `n_draws`:

```{r ex2-recert}
set.seed(9)
cert <- certify(res$projection, prob, n_draws = 5e5)
c(new_gap = cert$gap, new_se = cert$gap_se, new_gap_used = cert$gap_used,
  ess = cert$ess)
```

### When not to trust a certificate

`Var[G(theta)]` grows like `exp(|theta - a|^2)` in the distance from the nearest
atom, so a `theta` far from the support needs roughly `exp(|theta - a|^2)` draws
before its estimate means anything. A capable oracle will sometimes find such a
`theta` and report a large gap that is entirely sampling noise. The direction is
safe — the certificate only ever gets looser — but it can be vacuous, and **more
draws do not fix it**, because the radius the oracle can profitably exploit grows
with the sample too.

`certificate$ess` is the effective sample size behind the certified gap. Healthy
certificates report `ess` within an order of magnitude of `n_draws`;
noise-driven ones collapse to single digits, and `certify()` warns below
`ess_min`. Treat a warned certificate as
uninformative rather than as evidence of a large true gap — re-run it, or
certify a better-optimised projection.

Note that "sharper" does not mean "smaller": the inflation term halves, but the
raw maximum climbs toward the true `sup_theta G` for this (unpruned, imperfectly
optimised) mixture. More draws buys a more faithful bound, not automatically a
looser or tighter one. To actually lower `gap_used`, improve the projection.

`certify()` takes the fitted `projection` (or any `finite_mixing` of atoms on the
null) and sweeps the oracle over the null faces. It resamples the problem's own
engine by default; pass `engine =` to certify against a different one (say an
`exact_engine`).

`estimate = TRUE` draws a second independent sample and adds `gap_est`, an
*unbiased* point estimate of the gap at the selected `theta*`. It is a
diagnostic — useful for comparing `fw_variant`s or deciding whether to keep
optimising — and is biased downward relative to `sup_theta G`. It must never be
used to rescale an e-variable.
