---
output: github_document
---

<!-- README.md is generated from README.Rmd. Edit the .Rmd and run devtools::build_readme(). -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

# ripr

`ripr` computes the **reverse information projection** (RIPr) `P*` of an
alternative distribution `Q` onto a (possibly non-convex, union-structured)
null hypothesis, together with the **duality-gap certificate** for the rescaled
e-variable `(Q / P*) / (1 + gap)`. With an exact engine the rescaled e-variable's
expectation under every null distribution is at most 1. With a Monte Carlo engine
the gap is estimated, so the bound is stochastic: it holds at level `conf`
(default 0.95), at the worst-case `theta*` the certifying sample selects.

The projection is found by Frank--Wolfe and EM over a mixture of atoms on the
null faces. It requires the specification of a sampling `family` (the model
`p_theta`), the null geometry (a union of `face`s), and a surrogate `alternative`
`Q`. The alternative is an `outcome_distribution` — a law over the sample space —
which you can supply directly or build from a `mixing` measure over the
parameter space (e.g. `point_mixing`, `finite_mixing`), marginalised through the
family. These components are templated as
[S7](https://rconsortium.github.io/S7/) base classes, so new sample spaces drop
in without touching the algorithmic core. The algorithms require the ability to
compute KL divergences and gradients, which may be implemented either via
**exact** numerical methods or by **Monte Carlo** sampling. These two routes are
encapsulated in the `engine` class.


```{r install, eval = FALSE}
# install from github
renv::install("fleverest/ripr")
```


## Example 1: a one-sided binomial null

Let `X ~ Binomial(10, p)`, written as a 2-category multinomial with
`theta = (p, 1 - p)`. Consider testing `H_0: p <= 1/2` against `H_1: p > 1/2`,
adopting the simple alternative `Q = Binomial(10, 3/4)`. The RIPr of `Q` onto
the convex null is a **point mass at `p = 1/2`**. We demonstrate this using both
the exact and Monte Carlo engines.

```{r ex1, message = FALSE}
devtools::load_all()
n <- 10
fam <- multinomial_family(n_trials = n, k = 2)
# `ripr_problem` takes Q as an `outcome_distribution` over the sample space.
# Build one by marginalising a `mixing` measure over the parameter theta (here
# it is just a point mass at theta = (3/4, 1/4)) through the family, giving the
# outcome law Q = Binomial(10, 3/4).
Q <- as_marginal(point_mixing(theta_star = c(0.75, 0.25)), fam)

# H_0 = {p <= 1/2} is the simplex segment from p = 0 to the tie p = 1/2.
face <- polytope_face(vertices = cbind(c(0, 1), c(0.5, 0.5)), face_index = 1)
null <- null_region(faces = list(face))
```

**Exact engine** (enumerates all 11 outcomes):

```{r ex1-exact}
set.seed(1)
prob <- ripr_problem(fam, null, Q)
res <- run_ripr(
  prob,
  # Deliberately initialising somewhere sub-optimal for illustrative purposes
  init_atoms = matrix(c(0.25, 0.75), ncol = 1),
  init_atom_faces = 1L,
  fw_iters = 8,
  em_iters = 5,
  prune_threshold = 1e-6, # drop the numerically-dead atoms for output
  verbose = FALSE
)

cbind(
  p = res$projection@mixing@components[1, ],
  weight = res$projection@mixing@weights
)
```

`res$projection` is a `marginal` — the fitted `P*` as a distribution over
outcomes — and its `@mixing` is the `finite_mixing` holding the atoms and
weights on the null. `run_ripr()` returns the fitted e-variable directly; call
`e_value()` on new
data. Observing `(8, 2)` yields evidence against `H_0`, while a tie `(5, 5)`
yields evidence in favour of it:

```{r ex1-evalue}
res$e_variable
e_value(res$e_variable, rbind(c(8, 2), c(5, 5)))
```

**Monte Carlo engine** (1000 draws from `Q`) reaches the same projection:

```{r ex1-mc}
set.seed(3)
eng_mc <- mc_engine(fam, Q, n_draws = 1000)
prob_mc <- ripr_problem(fam, null, Q, engine = eng_mc)
res_mc <- run_ripr(
  prob_mc,
  init_atoms = matrix(c(0.25, 0.75), ncol = 1),
  init_atom_faces = 1L,
  fw_iters = 10,
  em_iters = 10,
  gap_tol = 1e-3,
  verbose = FALSE
)

cbind(
  p = res_mc$projection@mixing@components[1, ],
  weight = res_mc$projection@mixing@weights
)
```

`set.seed()` matters here: an `mc_engine` freezes its draw set at construction,
so a run is reproducible, but certification resamples and is not deterministic
across seeds.

## The result of a fit

`run_ripr()` returns six elements, partitioned by where each number came from:
`projection` and `e_variable` are the deliverables, `certificate` holds
everything measured on a fresh certification sample, `history` and `checkpoints`
hold everything measured on the fit sample, and `converged` reports whether the
*fit* gap met `gap_tol`.

```{r ex1-str}
names(res_mc)
```

The certificate reports the guaranteed e-value growth rate, the standard error
of the estimated gap, and the settings that produced it. Certification always
runs on draws independent of the fit, and on `10x` as many by default --
it is a single oracle sweep, not a per-iteration cost, so it is cheap to
certify far more thoroughly than you fit:

```{r ex1-cert}
unlist(res_mc$certificate[c(
  "gap", "gap_se", "gap_used", "growth_rate", "n_draws", "ess"
)])
```

`history` is one row per outer iteration, with the inner init/FW/EM steps and
the oracle argmax nested as list columns:

```{r ex1-history}
res_mc$history[, c("iter", "gap", "gap_se", "support_size", "kl_after_em")]
head(res_mc$history$kl_trace[[1]], 4) # the init + EM steps inside iteration 0
```

`checkpoints$final` is always present and describes the returned `projection`
*after* pruning. Its three fields are exactly the `init_atoms`,
`init_atom_faces` and `init_weights` arguments needed to resume the fit, so a
run can be continued without re-deriving anything (`init_weights` matters once a
projection has more than one atom — without it the atoms carry over but the
mixture restarts from uniform weights):

```{r ex1-final}
str(res_mc$checkpoints$final, max.level = 1)
```

### Re-certifying on a larger sample

The Monte Carlo certificate is a stochastic bound, and deliberately a
conservative one. `certify()` resamples fresh draws from `Q` (it does *not*
reuse the fit draws) and takes the oracle maximum over the null faces; that
maximum is biased *upward* for the true `sup_theta G`, which is the safe
direction, and `gap_used` inflates it further to a one-sided bound at level
`conf`. **`gap_used` is the only quantity that may be used to rescale an
e-variable.** Re-certifying the *same* fitted projection on more draws needs no
re-fitting — just raise `n_draws`:

```{r ex1-recert}
set.seed(9)
cert <- certify(res_mc$projection, prob_mc, n_draws = 1e5)
c(gap = cert$gap, se = cert$gap_se, gap_used = cert$gap_used, ess = cert$ess)
```

`certify()` takes the fitted `projection` (or any `finite_mixing` of atoms on the
null) and sweeps the oracle over the null faces. It resamples the problem's own
engine by default; pass `engine =` to certify against a different one (say an
`exact_engine`).

`estimate = TRUE` draws a second independent sample and adds `gap_est`, an
*unbiased* point estimate of the gap at the selected `theta*`. It is a
diagnostic — useful for comparing `fw_variant`s or deciding whether to keep
optimising — and is biased downward relative to `sup_theta G`. It must never be
used to rescale an e-variable.

### When not to trust a certificate

`Var[G(theta)]` grows like `exp(|theta - a|^2)` in the distance from the nearest
atom, so a `theta` far from the support needs roughly `exp(|theta - a|^2)` draws
before its estimate means anything. A capable oracle will sometimes find such a
`theta` and report a large gap that is entirely sampling noise. The direction is
safe — the certificate only ever gets looser — but it can be vacuous, and **more
draws do not fix it**, because the radius the oracle can profitably exploit grows
with the sample too.

`certificate$ess` is the effective sample size behind the certified gap. Healthy
certificates report `ess` within an order of magnitude of `n_draws` (above, the
whole sample); noise-driven ones collapse to single digits, and `certify()` warns
below `ess_min`. Treat a warned certificate as uninformative rather than as
evidence of a large true gap — re-run it, or certify a better-optimised
projection.

## Example 2

TODO: Gaussian plurality via quadrature/monte carlo.
