---
output: github_document
---

<!-- README.md is generated from README.Rmd. Edit the .Rmd and run devtools::build_readme(). -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

# ripr

`ripr` computes the **reverse information projection** (RIPr) `P*` of an
alternative distribution `Q` onto a (possibly non-convex, union-structured)
null hypothesis, together with the **duality-gap certificate** for the rescaled
e-variable `(Q / P*) / (1 + gap)`.

How strong that certificate is depends on how the gap was obtained, and every
certificate says which case it is in via `gap_certified`. For a multinomial
family on simplex-shaped null faces with an exact engine, pass
`bnb = bnb_control()` and the gap comes from a Bernstein branch-and-bound
**upper** bound on `sup_theta G`: the rescaled e-variable's expectation under
every null distribution is then at most 1 as a proven inequality. Otherwise the
gap comes from a multi-start BFGS search, which returns a *lower* bound on the
face maximum — so a missed global optimum leaves the e-variable
under-corrected, and the guarantee is evidence rather than proof. With a Monte
Carlo engine there is a second, separate error: the gap is estimated, so the
bound is stochastic and holds at level `conf` (default 0.95), at the worst-case
`theta*` the certifying sample selects.

The projection is found by Frank--Wolfe and EM over a mixture of atoms on the
null faces. It requires the specification of a sampling `family` (the model
`p_theta`), the null geometry (a union of `face`s), and a surrogate `alternative`
`Q`. The alternative is an `outcome_distribution` — a law over the sample space —
which you can supply directly or build from a `mixing` measure over the
parameter space (e.g. `point_mixing`, `finite_mixing`), marginalised through the
family. These components are templated as
[S7](https://rconsortium.github.io/S7/) base classes, so new sample spaces drop
in without touching the algorithmic core. The algorithms require the ability to
compute KL divergences and gradients, which may be implemented either via
**exact** numerical methods or by **Monte Carlo** sampling. These two routes are
encapsulated in the `engine` class.


```{r install, eval = FALSE}
# install from github
renv::install("fleverest/ripr")
```


## Example 1: a one-sided binomial null

Let `X ~ Binomial(10, p)`, written as a 2-category multinomial with
`theta = (p, 1 - p)`. Consider testing `H_0: p <= 1/2` against `H_1: p > 1/2`,
adopting the simple alternative `Q = Binomial(10, 3/4)`. The RIPr of `Q` onto
the convex null is a **point mass at `p = 1/2`**. We demonstrate this using both
the exact and Monte Carlo engines.

```{r ex1, message = FALSE}
library(ripr)
n <- 10
fam <- multinomial_family(n_trials = n, k = 2)
# `ripr_problem` takes Q as an `outcome_distribution` over the sample space.
# Build one by marginalising a `mixing` measure over the parameter theta (here
# it is just a point mass at theta = (3/4, 1/4)) through the family, giving the
# outcome law Q = Binomial(10, 3/4).
Q <- as_marginal(point_mixing(theta_star = c(0.75, 0.25)), fam)

# H_0 = {p <= 1/2} is the simplex segment from p = 0 to the tie p = 1/2.
face <- polytope_face(vertices = cbind(c(0, 1), c(0.5, 0.5)), face_index = 1)
null <- null_region(faces = list(face))
```

**Exact engine** (enumerates all 11 outcomes):

```{r ex1-exact}
set.seed(1)
prob <- ripr_problem(fam, null, Q)
res <- run_ripr(
  prob,
  # Deliberately initialising somewhere sub-optimal for illustrative purposes
  init_atoms = matrix(c(0.25, 0.75), ncol = 1),
  init_atom_faces = 1L,
  fw_iters = 8,
  em_iters = 5,
  prune_threshold = 1e-6, # drop the numerically-dead atoms for output
  verbose = FALSE
)

cbind(
  p = res$projection@mixing@components[1, ],
  weight = res$projection@mixing@weights
)
```

`res$projection` is a `marginal` — the fitted `P*` as a distribution over
outcomes — and its `@mixing` is the `finite_mixing` holding the atoms and
weights on the null. `run_ripr()` returns the fitted e-variable directly; call
`e_value()` on new
data. Observing `(8, 2)` yields evidence against `H_0`, while a tie `(5, 5)`
yields evidence in favour of it:

```{r ex1-evalue}
res$e_variable
```

```{r}
e_value(res$e_variable, rbind(c(8, 2), c(5, 5)))
```

### Certifying the gap deterministically

`gap` is what makes the rescaling valid, and by default it comes from a search:
`oracle()` maximises `G(theta) = E_Q[p_theta / P*]` by multi-start BFGS, so it
returns a *lower* bound on the face maximum. For a multinomial family on
simplex-shaped faces with an exact engine, the gap can instead be **proved**.
`p_theta(x)` is the degree-`n` Bernstein basis function at the count vector `x`,
so `G` already *is* a Bernstein form, with coefficients `q(x) / P*(x)`. That
basis is non-negative and sums to one, so the largest coefficient bounds `G`
across the whole face for free, and de Casteljau subdivision refines it.
`certifiable()` reports whether a problem qualifies:

```{r ex1-certifiable}
certifiable(face, prob$engine, fam)
```

```{r ex1-bnb}
heuristic <- certify(res$projection, prob, n_draws = 1e5)
certified <- certify(res$projection, prob, bnb = bnb_control())

data.frame(
  method = c(heuristic$gap_method, certified$gap_method),
  theta_star = c(heuristic$oracle_theta[1], certified$oracle_theta[1]),
  gap_used = c(heuristic$gap_used, certified$gap_used),
  certified = c(heuristic$gap_certified, certified$gap_certified)
)
```

The certified gap is the *larger* of the two, and that is the whole point. `G`
is increasing on this null, so its supremum sits at the face's boundary vertex
`p = 1/2` — but `oracle()` searches in a softmax parametrisation that covers
only the face's *interior*, and the objective is nearly flat there, so BFGS
stops just short, at the projection's own atom. The bound evaluates the vertex
exactly.

Small as the difference is (in this example), it is the difference between an
e-variable and not one. `E_theta[e(X)]` must be at most 1 at every `theta` inthe null:

```{r ex1-guarantee}
X <- support(fam)
null_expectation <- function(gap, p) {
  ev <- e_variable(numerator = Q, projection = res$projection, gap = gap)
  sum(exp(log_density(fam, c(p, 1 - p)) + e_value(ev, X, log = TRUE)))
}
# Distance above 1 at the worst-case theta. Positive is a violation.
c(
  heuristic = null_expectation(heuristic$gap_used, 0.5) - 1,
  certified = null_expectation(certified$gap_used, 0.5) - 1
)
```

`bnb_iterations` is 0 here: the largest Bernstein coefficient already sits at
the vertex attaining the maximum; with a single atom mixture, the likelihood
ratio in monotone in `p`. The bound is exactly tight before any subdivision
happens. Harder faces bisect until `bound - incumbent <= tol`. The bound is
valid at *every* iteration, so `tol` and `max_iter` buy tightness, never
validity.

To carry the guarantee through a fit rather than re-certifying afterwards, pass
`certify_bnb` — the returned `e_variable` then records the guarantee level, so a
caller holding only that object can still tell a proven rescaling from a
heuristic one:

```{r ex1-run-certified}
set.seed(1)
res_cert <- run_ripr(
  prob,
  init_atoms = matrix(c(0.25, 0.75), ncol = 1),
  init_atom_faces = 1L,
  fw_iters = 8,
  em_iters = 5,
  prune_threshold = 1e-6,
  verbose = FALSE,
  certify_bnb = bnb_control()
)
res_cert$e_variable
```

This path needs all three of an exact engine, a family whose density is a
Bernstein basis function of `theta` (the multinomial's is; a Gaussian's is not),
and bounded simplex faces — `halfspace_face` has no simplex to subdivide, which
is a permanent gap rather than a missing feature. Every other combination
degrades to the heuristic path and reports `gap_certified = FALSE` rather than
pretending; `require_certified = TRUE` turns that fallback into an error naming
what blocked it. See `?certify` for what the two regimes cover, and
`bnb_control()`'s `max_coef` for the cost ceiling — coefficients per sub-simplex
are `choose(n + K - 1, K - 1)`, so certification caps the usable batch size more
tightly than optimisation does.

**Monte Carlo engine** (1000 draws from `Q`) reaches the same projection:

```{r ex1-mc}
set.seed(3)
eng_mc <- mc_engine(fam, Q, n_draws = 1000)
prob_mc <- ripr_problem(fam, null, Q, engine = eng_mc)
res_mc <- run_ripr(
  prob_mc,
  init_atoms = matrix(c(0.25, 0.75), ncol = 1),
  init_atom_faces = 1L,
  fw_iters = 10,
  em_iters = 10,
  gap_tol = 1e-3,
  verbose = FALSE
)

cbind(
  p = res_mc$projection@mixing@components[1, ],
  weight = res_mc$projection@mixing@weights
)
```

`set.seed()` matters here: an `mc_engine` freezes its draw set at construction,
so a run is reproducible, but certification resamples and is not deterministic
across seeds.

## The result of a fit

`run_ripr()` returns six elements, partitioned by where each number came from:
`projection` and `e_variable` are the deliverables, `certificate` holds
everything measured on a fresh certification sample, `history` and `checkpoints`
hold everything measured on the fit sample, and `converged` reports whether the
*fit* gap met `gap_tol`.

```{r ex1-str}
names(res_mc)
```

The certificate reports the guaranteed e-value growth rate, the standard error
of the estimated gap, and the settings that produced it. Certification always
runs on draws independent of the fit, and on `10x` as many by default --
it is a single oracle sweep, not a per-iteration cost, so it is cheap to
certify far more thoroughly than you fit:

```{r ex1-cert}
unlist(res_mc$certificate[c(
  "gap",
  "gap_se",
  "gap_used",
  "growth_rate",
  "n_draws",
  "ess"
)])
```

`history` is one row per outer iteration, with the inner init/FW/EM steps and
the oracle argmax nested as list columns:

```{r ex1-history}
res_mc$history[, c("iter", "gap", "gap_se", "support_size", "kl_after_em")]
head(res_mc$history$kl_trace[[1]], 4) # the init + EM steps inside iteration 0
```

`checkpoints$final` is always present and describes the returned `projection`
*after* pruning. Its three fields are exactly the `init_atoms`,
`init_atom_faces` and `init_weights` arguments needed to resume the fit, so a
run can be continued without re-deriving anything (`init_weights` matters once a
projection has more than one atom — without it the atoms carry over but the
mixture restarts from uniform weights):

```{r ex1-final}
str(res_mc$checkpoints$final, max.level = 1)
```

### Re-certifying on a larger sample

The Monte Carlo certificate is a stochastic bound, and deliberately a
conservative one. `certify()` resamples fresh draws from `Q` (it does *not*
reuse the fit draws) and takes the oracle maximum over the null faces; that
maximum is biased *upward* for the true `sup_theta G`, which is the safe
direction, and `gap_used` inflates it further to a one-sided bound at level
`conf`. **`gap_used` is the only quantity that may be used to rescale an
e-variable.** Re-certifying the *same* fitted projection on more draws needs no
re-fitting — just raise `n_draws`:

```{r ex1-recert}
set.seed(9)
cert <- certify(res_mc$projection, prob_mc, n_draws = 1e5)
c(gap = cert$gap, se = cert$gap_se, gap_used = cert$gap_used, ess = cert$ess)
```

`certify()` takes the fitted `projection` (or any `finite_mixing` of atoms on the
null) and sweeps the oracle over the null faces. It resamples the problem's own
engine by default; pass `engine =` to certify against a different one (say an
`exact_engine`).

`estimate = TRUE` draws a second independent sample and adds `gap_est`, an
*unbiased* point estimate of the gap at the selected `theta*`. It is a
diagnostic — useful for comparing `fw_variant`s or deciding whether to keep
optimising — and is biased downward relative to `sup_theta G`. It must never be
used to rescale an e-variable.

## Example 2: A 4-category plurality null

Let `X ~ Multinomial(25, theta)`. Consider testing `H_0: theta_1 <= theta_i` for
some `i!=1` against `H_1: theta_1 > theta_i` for all `i!=1`, adopting the simple
alternative `Q = Multinomial(25, (0.3, 0.27, 0.23, 0.15, 0.05))`. The RIPr of `Q` onto
the null is no longer a point mass on the boundary.

```{r ex2, message = FALSE}
n <- 25
k <- 5
q <- c(0.3, 0.27, 0.23, 0.15, 0.05)

# The plurality null can be written as H_0 = H_02 ∪ H_03 ∪ H_04, where
# H_0i = {theta_1 <= theta_i} is a subsimplex with vertices satisfying
# (i) theta_1=theta_i=1/2, (ii) theta_1=theta_i=0, and
# (iii) theta_j = 1 for all other j.
plurality_faces <- function(K) {
  lapply(2:K, function(j) {
    basis <- lapply(setdiff(seq_len(K), 1L), function(k) {
      v <- numeric(K)
      v[k] <- 1
      v
    })
    tie <- numeric(K)
    tie[c(1L, j)] <- 0.5
    polytope_face(
      vertices = do.call(cbind, c(basis, list(tie))),
      face_index = j
    )
  })
}

fam <- multinomial_family(n_trials = n, k = k)
Q <- as_marginal(point_mixing(theta_star = q), fam)
faces <- plurality_faces(k)
null <- null_region(faces = faces)
prob <- ripr_problem(fam, null, Q)

# atoms to start from: one for each face
init <- do.call(
  cbind,
  lapply(faces, function(f) init_point(f, q))
)

res <- run_ripr(
  prob,
  init_atoms = init,
  init_atom_faces = seq_len(k - 1),
  fw_iters = 10,
  em_iters = 10,
  verbose = FALSE
)

heuristic <- certify(res$projection, prob, n_draws = 1e5)
certified <- certify(res$projection, prob, bnb = bnb_control(tol = 1e-9))

data.frame(
  method = c(heuristic$gap_method, certified$gap_method),
  gap_used = c(heuristic$gap_used, certified$gap_used),
  certified = c(heuristic$gap_certified, certified$gap_certified)
)
```


## Example 3

TODO: Gaussian plurality via quadrature/monte carlo.
