---
output: github_document
bibliography: inst/REFERENCES.bib
link-citations: true
---

<!-- README.md is generated from README.Rmd. Edit the .Rmd and run devtools::build_readme(). -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  dev = "svg",
  out.width = "100%"
)
```

```{r helpers, include = FALSE}
tri_xy <- function(theta) {
  rbind(c(0.5, 0, 1), c(sqrt(3) / 2, 0, 0)) %*% as.matrix(theta)
}

plot_null <- function(
  null,
  points = NULL,
  weights = NULL,
  main = NULL,
  fill = "#c96f5b",
  bg = c("#2c6f9b", "white"),
  alpha = 0.55
) {
  corners <- tri_xy(diag(3))
  op <- par(mar = c(0.5, 0.5, if (is.null(main)) 0.5 else 2, 0.5))
  on.exit(par(op), add = TRUE)
  plot.new()
  plot.window(c(-0.05, 1.05), c(-0.05, 0.92), asp = 1)
  for (s in null@subnulls) {
    xy <- tri_xy(s@vertices)
    polygon(
      xy[1, ],
      xy[2, ],
      col = adjustcolor(fill, 0.45),
      border = fill,
      lwd = 1.5
    )
  }
  polygon(corners[1, ], corners[2, ], border = "grey50")
  text(
    corners[1, ],
    corners[2, ],
    c(1, 2, 3),
    pos = c(3, 1, 1),
    cex = 0.8,
    col = "grey40"
  )
  if (!is.null(points)) {
    groups <- if (is.list(points)) points else list(points)
    sizes <- if (is.list(weights)) weights else list(weights)
    # Marker area proportional to mixing weight -- `cex` scales a linear
    # dimension, so the honest map is sqrt -- on a scale shared across groups,
    # so a heavy atom looks heavy wherever it sits.
    all_w <- unlist(sizes)
    for (g in seq_along(groups)) {
      xy <- tri_xy(groups[[g]])
      w <- sizes[[min(g, length(sizes))]]
      cex <- if (is.null(w)) 1.2 else pmax(0.35, 1.7 * sqrt(w / max(all_w)))
      points(
        xy[1, ],
        xy[2, ],
        pch = 21,
        bg = adjustcolor(bg[[(g - 1L) %% length(bg) + 1L]], alpha),
        col = "grey15",
        cex = cex,
        lwd = 1.1
      )
    }
  }
  if (!is.null(main)) {
    title(main, font.main = 1, cex.main = 1)
  }
  invisible(NULL)
}
```

# ripr

`ripr` computes reverse information projections onto nulls that take the form of a union of convex parts, and then proves an upper bound on the expectation over the null of whatever random variable you get.
The projection is fitted via Li--Barron [@LiBarron1999], Frank--Wolfe [@Jaggi2013] and/or EM.
Bounds on the null expectation are proved by a branch and bound algorithm (currently only for multinomial via the Bernstein enclosure).
At the end, you get a random variable that is genuinely an e-variable for the respective null.

The idea it rests on is simpler than the reverse information projection itself. For any non-negative, bounded random variable `X`, and `b` is any number at least `sup_{theta in Theta_0} E_theta[X]`, then `X / b` has null expectation at most one; i.e. `E = X / b` is an e-variable.
Nothing requires `X` to be the true minimiser of KL divergence.
**This means that we need not fit the projection exactly:** any approximation can be used so long as we correct the resulting likelihood ratio to yield a valid e-variable, and the quality of the approximation determines only how much e-power survives the correction.
That splits a hard, non-convex optimisation problem into an easy one (approximate) and a proof (bound), and it this is what `ripr` facilitates.


## Installation

```{r install, eval = FALSE}
# install.packages("remotes")
remotes::install_github("fleverest/ripr")
```

## Approximate, then correct

Take a three-category multinomial distribution (with `n = 20`) and the plurality null: candidate 1 does not win outright, `H0 = union_j {theta : theta_1 <= theta_j}`.
This null can be written as a union of convex parts; one region where `theta_1 <= theta_2` and another where `theta_1 <= theta_3`.
Each part (or sub-null) is a simplex, so we represent the plurality null as a single `null_model` holding two `simplex_null` subnulls.

```{r problem, message = FALSE}
library(ripr)

K <- 3L
n <- 20L
family <- multinomial_family(n_trials = n, k = K)

plurality <- null_model(
  family,
  lapply(2:K, function(j) {
    # The two sub-simplices: one for each subnull.
    vertices <- diag(K)
    vertices[, 1L] <- replace(numeric(K), c(1L, j), 0.5)
    simplex_null(vertices = vertices)
  })
)
```

We nominate a discrete mixture-multinomial distribution, in this case the mixing measure is just a single dirac at `q = (0.40, 0.35, 0.25)`, yielding a simple `Multinomial(20, q)` alternative.

```{r alternative}
W1 <- finite_mixing(
  components = cbind(
    c(0.40, 0.35, 0.25)
  ),
  weights = c(1)
)
Q <- mixture(W1, family)
```

```{r plot-plurality}
#| echo: false
#| fig.width: 4.5
#| fig.height: 4
#| out.width: "60%"
#| fig.align: "center" 
#| fig.cap: Plurality null, with the support of the alternative mixing distribution W1.
plot_null(
  plurality,
  points = W1@components,
  weights = W1@weights / 4,
  bg = "white"
)
```

The two subnulls overlap, and the alternative sits in the cap above them where candidate 1 leads.
The fitting procedure is implemented as a sequence of steps: `fw_step` for Frank--Wolfe updates and `em_step` for EM (also `lb_step` for Li--Barron, but not recommended).
Here we demonstrate running 40 steps of a Frank--Wolfe procedure.
After running Frank--Wolfe, `ripr_finish` can be used to solve the weights, and remove any low-mass atoms.

```{r fit}
set.seed(1)
state <- ripr_init(Q, plurality)
state <- fw_step(state, times = 40L, until = gap_below(1e-10))
fit <- ripr_finish(state, reoptimise = TRUE, identify = TRUE, record_gap = TRUE)

c(kl = fit$kl, gap = fit$gap_final, atoms = ncol(fit$W0@components))
```

```{r plot-plurality-fit}
#| echo: false
#| fig.width: 4.5
#| fig.height: 4
#| out.width: "60%"
#| fig.align: "center" 
#| fig.cap: Plurality null, with the support of the mixing distributions W1 (alternative) and W0 (null, fitted).
plot_null(
  plurality,
  points = list(fit$W0@components, W1@components),
  weights = list(fit$W0@weights, W1@weights / 4)
)
```

The fit projection attains a KL divergence (from `Q`) of approximately `r round(fit$kl, 4)`.
The `gap` quanitty corresponds to an estimate of the Frank--Wolfe duality gap, which at the true value bounds the expectation of the ratio `Q / P*` over the null by `1 + gap`.
However, our `gap` estimate here is not an upper bound on the true quantity, so it can not be used directly to obtain a certified e-value.

With our fit projection, we may construct a `random_variable` that computes the mixture likelihood ratio `Q / P*`, i.e. our candidate e-variable:

```{r ratio}
X <- mixture_likelihood(Q, label = "Q") /
  mixture_likelihood(fit$P_star, label = "P*")
X
```

`random_variable`s are effectively functions that map elements of the sample space to their realisation:

```{r rv_outcome}
outcomes <- rbind(
  c(10L, 10L, 0L),
  c(8L, 7L, 5L)
)
X(outcomes)
```

Finally, we may use `certify()` to prove an upper bound on the null expectation.
It returns a proven upper bound alongside the largest value the search actually attained, and the two bracket the true upper bound which is generally unknown:

```{r certify}
cert <- certify(X, plurality, tol = 1e-9)
c(
  upper = cert$sup_ub,
  attained = cert$sup_lb,
  width = cert$sup_ub - cert$sup_lb
)
```

The bound lands just above one: it precisely bounds `1 + gap`, with `gap` the *true* Frank--Wolfe duality gap that the fit stopped on.

```{r identity}
c(bound_minus_one = cert$sup_ub - 1, fitted_gap = fit$gap_final)
```

The re-scaled random variable `E = X / cert$sup_ub` is then a genuine e-variable for `plurality`:

```{r evariable}
E <- X / cert$sup_ub
print(E)
print(E(outcomes))
```


### The correction does not need a good fit

Replace the projection with something that is not the RIPr at all and certify the resulting ratio:

```{r bad-fit}
bad <- mixture(point_mixing(rep(1 / K, K)), family)
X_bad <- mixture_likelihood(Q, label = "Q") /
  mixture_likelihood(bad, label = "P0")
cert_bad <- certify(X_bad, plurality, tol = 1e-9)

c(good = cert$sup_ub, bad = cert_bad$sup_ub)
```

Both rescale to valid e-variables.
The important quality that differentiates them is e-power, i.e. log-growth: the expected log e-value under the alternative, which is what compounds over a sequential procedure.
We could measure growth as follows.

```{r growth}
growth <- function(E, Q) {
  outcomes <- enumerate_space(E@sample_space)
  e_values <- E(outcomes)
  Q_expectation <- dist_log_density(Q, outcomes) |>
    exp() %*%
    log(e_values)
  c(Q_expectation) # return as scalar
}

E_gr <- growth(E, Q)

c(good = E_gr, bad = growth(X_bad / cert_bad$sup_ub, Q))
```

A negative growth rate means that we lose evidence on average even when the alternative is true: useless, but not invalid.
**Validity is free from the certification and correction procedure; while fitting a good approximation buys e-power.** 

### How close to optimal is `E`?

When `P*` is the true reverse information projection of `Q` onto `null`, the attained KL divergence `KL(Q || P*)` is precisely the growth-rate of the resulting e-variable `Q / P*`.
For any fitted `P`, the KL divergence is above the optimum, so the final kl divergence of the fit upper bounds the optimal KL, and thus the optimal growth-rate.
Our e-variable will not be growth-rate optimal unless it is the true minimiser.
Therefore we know the optimum growth-rate will lie somewhere in the interval between `growth(E, Q)` and `fit$kl` (provided `fit$kl` is not an estimate, e.g. via Monte Carlo):

```{r growth-kl-gap}
c(
  lower = E_gr,
  upper = fit$kl,
  width = fit$kl - E_gr
)
```


## A different multinomial null

Consider the null `H0: theta_1, theta_2, theta_3 <= 1 / 2`, i.e. no category holds an outright majority (exceeds a proportion of `1 / 2`).
This forms the medial triangle in the 2-simplex.
We nominate a discrete mixture alternative with one dirac per section in the alternative, equally weighted.

```{r medial}
medial <- null_model(
  family,
  list(simplex_null(
    vertices = cbind(c(0.5, 0.5, 0), c(0.5, 0, 0.5), c(0, 0.5, 0.5))
  ))
)

W_maj <- finite_mixing(
  components = cbind(
    c(0.6, 0.2, 0.2),
    c(0.2, 0.6, 0.2),
    c(0.2, 0.2, 0.6)
  ),
  weights = c(1 / 3, 1 / 3, 1 / 3)
)
Q_maj <- mixture(W_maj, family)
```

```{r plot-medial-null}
#| echo: false
#| fig.width: 4.5
#| fig.height: 4
#| out.width: "60%"
#| fig.align: "center" 
#| fig.cap: Medial-triangle null, with the support of W_maj.
plot_null(
  medial,
  points = W_maj@components,
  weights = W_maj@weights / 4,
  bg = "white"
)
```

Here we demonstrate a slightly more elegant fitting procedure here that interleaves Frank--Wolfe and EM steps:

```{r medial-alternative}
hybrid_step <- function(state, times) {
  for (iter in seq_len(times)) {
    state <- state |> fw_step() |> em_step()
  }
  state
}

set.seed(2)
fit_medial <- ripr_init(Q_maj, medial) |>
  hybrid_step(times = 40L) |>
  ripr_finish(
    reoptimise = TRUE,
    identify = TRUE,
    record_gap = TRUE
  )
X_medial <- mixture_likelihood(Q_maj, label = "Q") /
  mixture_likelihood(fit_medial$P_star, label = "P*")

c(kl = fit_medial$kl, upper = certify(X_medial, medial, tol = 1e-9)$sup_ub)
```

```{r plot-medial-fit}
#| echo: false
#| fig.width: 4.5
#| fig.height: 4
#| out.width: "60%"
#| fig.align: "center" 
#| fig.cap: Medial-triangle null, with the support of W_maj and W0 (fitted).
plot_null(
  medial,
  points = list(fit_medial$W0@components, W_maj@components),
  weights = list(fit_medial$W0@weights, W_maj@weights / 4)
)
```

## Searching for a supremum is not the same as proving a bound

`sup_lb()` hunts for the largest null expectation by multi-start gradient ascent.
It is much cheaper than proving a bound, and it works on families and geometries that may not have an implemented bounding procedure, but it is **not** a certificate.
It reports what it found; a lower bound on the supremum.
Something larger may sit somewhere difficult to reach via gradient ascent.

```{r sup-lb}
c(searched = sup_lb(X, plurality)$sup_lb, certified = cert$sup_ub)
```

Where no bounding method has been implemented, `certify()` refuses:

```{r refuse, error = TRUE}
gaussian <- gaussian_family(dim = 2L)
null <- null_model(
  gaussian,
  list(halfspace_null(normal = c(1, -1), offset = 0))
)
Y <- mixture_likelihood(mixture(point_mixing(c(0, 0)), gaussian))
certify(Y, null)
```


## References

`ripr` targets the growth-rate optimal e-variable of safe testing [@GrunwaldDeHeideKoolen2024], approximating its reverse information projection via the numeraire characterisation of @LarssonRamdasRuf2025.
Certified bounds use the branch-and-bound scheme of @Leroy2012, the simplicial Bernstein range enclosure of @Garloff1986, and the subdivision techniques of @PrautzschBoehmPaluszny2002 (chapters 10--11).
