---
title: "Regions and set algebra"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Regions and set algebra}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 4.5,
  fig.height = 4,
  out.width = "60%",
  fig.align = "center",
  dev = "svg"
)
```

```{r helpers, include = FALSE}
tri_xy <- function(theta) {
  rbind(c(0.5, 0, 1), c(sqrt(3) / 2, 0, 0)) %*% as.matrix(theta)
}

plot_null <- function(
  null,
  points = NULL,
  weights = NULL,
  main = NULL,
  fill = "#c96f5b",
  bg = c("#2c6f9b", "white"),
  alpha = 0.55
) {
  corners <- tri_xy(diag(3))
  op <- par(mar = c(0.5, 0.5, if (is.null(main)) 0.5 else 2, 0.5))
  on.exit(par(op), add = TRUE)
  plot.new()
  plot.window(c(-0.05, 1.05), c(-0.05, 0.92), asp = 1)
  for (s in parts(null@region)) {
    xy <- tri_xy(s@vertices)
    polygon(
      xy[1, ],
      xy[2, ],
      col = adjustcolor(fill, 0.45),
      border = fill,
      lwd = 1.5
    )
  }
  polygon(corners[1, ], corners[2, ], border = "grey50")
  text(
    corners[1, ],
    corners[2, ],
    c(1, 2, 3),
    pos = c(3, 1, 1),
    cex = 0.8,
    col = "grey40"
  )
  if (!is.null(points)) {
    groups <- if (is.list(points)) points else list(points)
    sizes <- if (is.list(weights)) weights else list(weights)
    all_w <- unlist(sizes)
    for (g in seq_along(groups)) {
      xy <- tri_xy(groups[[g]])
      w <- sizes[[min(g, length(sizes))]]
      cex <- if (is.null(w)) 1.2 else pmax(0.35, 1.7 * sqrt(w / max(all_w)))
      points(
        xy[1, ],
        xy[2, ],
        pch = 21,
        bg = adjustcolor(bg[[(g - 1L) %% length(bg) + 1L]], alpha),
        col = "grey15",
        cex = cex,
        lwd = 1.1
      )
    }
  }
  if (!is.null(main)) {
    title(main, font.main = 1, cex.main = 1)
  }
  invisible(NULL)
}
```

```{r library, message = FALSE}
library(ripr)
```

## Introduction

A null hypothesis, or `null_model` in `ripr` comprises a `parametric_family` and a `region`: a subset of the family's `parameter_space`. `region` is abstract and splits in two.
`convex_region`s are individual convex units that carry the geometry a fitting procedure needs: a `chart()` provides a coordinate system for optimisation, `contains()` encodes membership for the set, and `project()` moves outside points onto the region.
The `union_region` is a finite union of convex regions and is generally *not* convex.

```{r hierarchy}
s <- simplex_region(vertices = diag(3))
S7::S7_inherits(s, region)
S7::S7_inherits(s, convex_region)

u <- union_region(s, halfspace_region(normal = c(1, -1, 0)))
S7::S7_inherits(u, region)
S7::S7_inherits(u, convex_region)
```

There are two methods that all unions support: `parts()` and `cells()`.
`parts()` returns all the convex pieces that the region was *declared* as, whereas `cells()` returns a triangulation.
The two agree for a `union_region`, but not for a `polytope_region` that isn't already a simplex: the certification routine needs simplices while fitting can operate directly on polytopes, so a square region has one part and two triangular cells, each filling different roles in the package.

```{r cells-vs-parts}
square <- polytope_region(vertices = cbind(c(0, 0), c(1, 0), c(1, 1), c(0, 1)))
n_parts(square)
n_cells(square)
```

## Convex geometries

Every convex region in the package is, underneath, a `polyhedron_region()`: the Minkowski--Weyl generator form `conv(V) + cone(R) + span(L)` (or its dual facet description, if you build one from `h_region()` instead).
`simplex_region()`, `polytope_region()`, `point_region()`, `unconstrained_region()` and `halfspace_region()` special cases of convex polyhedra, we provide them as convenient constructors for more familiar objects, with validation.
`simplex_region()` and `halfspace_region()` are the two you've already seen in `vignette("ripr")`. Here are the rest:

```{r geometries}
# The convex hull of a vertex set. A simplex_region is the special case when
# the vertices are affinely independent.
polytope_region(vertices = cbind(c(0, 0), c(1, 0), c(1, 1), c(0, 1)))

# A single parameter point, e.g. to specify a simple null theta = (0.5 0.3 0.2)'
point_region(theta = c(0.5, 0.3, 0.2))

# All of R^2
unconstrained_region(2L)
```

## Set algebra

`region`s can be composes with `union()`, `intersect()`, `setdiff()`, `setequal()` and `disjoin()`, all working in exact rational arithmetic so overlaps and boundaries are decided precisely rather than approximately.
The plurality null from `vignette("ripr")` was constructed from two overlapping triangles.

```{r plurality-setup}
K <- 3L
plurality_part <- function(j) {
  vertices <- diag(K)
  vertices[, 1L] <- replace(numeric(K), c(1L, j), 0.5)
  simplex_region(vertices = vertices)
}
plurality <- union(plurality_part(2), plurality_part(3))
plurality
```

Calling `intersect()` instead yields the region where the two parts overlap: the wedge where `theta_1 <= theta_2, theta_3`.

```{r intersect-verb}
intersect(plurality_part(2), plurality_part(3))
```

`setdiff(x, y)` is the closure of `x` minus `y`.
The closed complement of the plurality null within the full simplex is the region where `theta_1 >= theta_2, theta_3`.
It is not quite the alternative because it includes the boundary where `theta_1 = theta_2` or `theta_1 = theta_3`.

```{r setdiff-verb}
leader <- setdiff(simplex_region(vertices = diag(K)), plurality)
leader
```

The plurality null's two parts overlap, which is fine for fitting and certification (a supremum over a union is a supremum over any cover of it), but sometimes you may wish for a disjoint cover instead.
For instance, to sum a measure over the null without double-counting the overlap.
`disjoin()` peels a union apart sequentially: it leaves the first part as is, then subtracts the first part from the second, and so on.

```{r disjoin-verb}
peeled <- disjoin(plurality)
n_parts(plurality)
n_parts(peeled)
plurality[2]
peeled[2]
```

`setequal()` decides whether two regions describe the same set.
This is computed exactly, regardless of how each set is decomposed into parts.

```{r setequal-verb}
setequal(peeled, plurality)
setequal(square, union_region(cells(square)))
setequal(square, cells(square)[[1]])
```

When nothing is left, the algebra says so rather than erroring out.
`intersect()` of disjoint regions, or a `setdiff()` that removes everything, is an `empty_region()`: it behaves like a list of length zero, and `union()` drops it without a trace.

```{r empty-region}
nothing <- intersect(
  point_region(theta = c(1, 0, 0)),
  point_region(theta = c(0, 1, 0))
)
is_empty(nothing)
identical(union(nothing, s), s)
```

## A Sierpinski-like null

`vignette("ripr")` builds the *medial* null for a three-category multinomial: the null region is where no candidate holds a majority, `H0: theta_1, theta_2, theta_3 <= 1/2`.
Geometrically it is the medial triangle of the 2-simplex, the one you get by joining the midpoints of each edge. The alternative is the union of the three corner triangles.

The corners themselves are easily obtained by taking the complement of the null within the full simplex via `setdiff()`:

```{r medial-corners}
medial_of <- function(triangle) {
  v <- triangle@vertices
  simplex_region(
    vertices = cbind(
      (v[, 1] + v[, 2]) / 2,
      (v[, 1] + v[, 3]) / 2,
      (v[, 2] + v[, 3]) / 2
    )
  )
}

full <- simplex_region(vertices = diag(3))
medial <- medial_of(full)
corners <- parts(setdiff(full, medial))
length(corners)
```

Repeating the same split inside each corner (take its own medial triangle, add that to the null) a few times yields a Sierpinski approximation built entirely from `union()` and `setdiff()`:

```{r sierpinski-null}
sierpinski_null <- function(triangle, depth) {
  medial <- medial_of(triangle)
  if (depth <= 1L) {
    return(medial)
  }
  corners <- parts(setdiff(triangle, medial))
  union(medial, lapply(corners, sierpinski_null, depth = depth - 1L))
}

depth <- 3L
geometry <- sierpinski_null(full, depth)
n_parts(geometry)
```

The finest, still-uncovered corners are the natural place to put an
alternative -- one dirac per leftover triangle, exactly as `vignette("ripr")`
puts one per corner of the plain medial null, just at finer resolution:

```{r sierpinski-leaves}
leaf_corners <- function(triangle, depth) {
  if (depth <= 0L) {
    return(list(triangle))
  }
  corners <- parts(setdiff(triangle, medial_of(triangle)))
  unlist(lapply(corners, leaf_corners, depth = depth - 1L), recursive = FALSE)
}

leaves <- leaf_corners(full, depth)
length(leaves)

centroids <- vapply(leaves, \(cell) rowMeans(cell@vertices), numeric(3))
W_alt <- finite_mixing(
  components = centroids,
  weights = rep(1 / length(leaves), length(leaves))
)
```

```{r plot-sierpinski-null, echo = FALSE, fig.cap = "A depth-3 Sierpinski-like null, with the support of the alternative -- one dirac per leftover corner."}
family <- multinomial_family(n_trials = 20L, k = 3L)
sierpinski <- null_model(family, geometry)
Q_alt <- family(W_alt)
plot_null(
  sierpinski,
  points = atoms(Q_alt@mixing),
  weights = weights(Q_alt@mixing) / 4,
  bg = "white"
)
```

Every part of `geometry` is still just a `simplex_region`, so nothing about
fitting or certification needs to know it came from a recursive construction
rather than being typed out by hand as two parts:

```{r sierpinski-fit}
hybrid_step <- function(state, times) {
  for (iter in seq_len(times)) {
    state <- state |> fw_step() |> em_step()
  }
  state
}

set.seed(3)
fit <- ripr_init(Q_alt, sierpinski) |>
  hybrid_step(times = 40L) |>
  ripr_finish(reoptimise = TRUE, identify = TRUE, record_gap = TRUE)

c(kl = fit$kl, atoms = n_atoms(fit$W0))
```

```{r plot-sierpinski-fit, echo = FALSE, fig.cap = "The same null, with the support of the alternative and the fitted null mixture W0."}
plot_null(
  sierpinski,
  points = list(atoms(fit$W0), atoms(Q_alt@mixing)),
  weights = list(weights(fit$W0), weights(Q_alt@mixing) / 4)
)
```

```{r sierpinski-certify}
X <- likelihood(Q_alt, label = "Q") /
  likelihood(fit$P_star, label = "P*")
cert <- certify(X, sierpinski, tol = 1e-9)
c(upper = cert$sup_ub, attained = cert$sup_lb)
```

`certify()` proves the same kind of bound it did for the two-part plurality null in `vignette("ripr")`, over a `union_region` of `r n_parts(geometry)` parts built by a few lines of recursive set algebra rather than declared by hand.
`E <- X / cert$sup_ub` is a genuine e-variable for `sierpinski`, exactly as before.
The gap is quite large, however, which indicates that it is probably quite a poor fit.

## Where next

`vignette("ripr")` covers the other half: fitting a projection, why validity survives a deliberately bad one, and what `certify()` refuses to do without a bounding method.
If you're implementing a new convex geometry rather than composing the built-in ones, `?convex_region` and its generics (`chart()`, `contains()`, `project()`, `init_point()`) describe the interface it must satisfy.
