---
title: "Getting Started with ripr"
output: rmarkdown::html_vignette
bibliography: ../inst/REFERENCES.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Getting Started with ripr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 4.5,
  fig.height = 4,
  out.width = "60%",
  fig.align = "center",
  dev = "svg"
)
```

```{r helpers, include = FALSE}
tri_xy <- function(theta) {
  rbind(c(0.5, 0, 1), c(sqrt(3) / 2, 0, 0)) %*% as.matrix(theta)
}

plot_null <- function(
  null,
  points = NULL,
  weights = NULL,
  main = NULL,
  fill = "#c96f5b",
  bg = c("#2c6f9b", "white"),
  alpha = 0.55
) {
  corners <- tri_xy(diag(3))
  op <- par(mar = c(0.5, 0.5, if (is.null(main)) 0.5 else 2, 0.5))
  on.exit(par(op), add = TRUE)
  plot.new()
  plot.window(c(-0.05, 1.05), c(-0.05, 0.92), asp = 1)
  for (s in parts(null@region)) {
    xy <- tri_xy(s@vertices)
    polygon(
      xy[1, ],
      xy[2, ],
      col = adjustcolor(fill, 0.45),
      border = fill,
      lwd = 1.5
    )
  }
  polygon(corners[1, ], corners[2, ], border = "grey50")
  text(
    corners[1, ],
    corners[2, ],
    c(1, 2, 3),
    pos = c(3, 1, 1),
    cex = 0.8,
    col = "grey40"
  )
  if (!is.null(points)) {
    groups <- if (is.list(points)) points else list(points)
    sizes <- if (is.list(weights)) weights else list(weights)
    all_w <- unlist(sizes)
    for (g in seq_along(groups)) {
      xy <- tri_xy(groups[[g]])
      w <- sizes[[min(g, length(sizes))]]
      cex <- if (is.null(w)) 1.2 else pmax(0.35, 1.7 * sqrt(w / max(all_w)))
      points(
        xy[1, ],
        xy[2, ],
        pch = 21,
        bg = adjustcolor(bg[[(g - 1L) %% length(bg) + 1L]], alpha),
        col = "grey15",
        cex = cex,
        lwd = 1.1
      )
    }
  }
  if (!is.null(main)) {
    title(main, font.main = 1, cex.main = 1)
  }
  invisible(NULL)
}
```

```{r library, message = FALSE}
library(ripr)
```

## Introduction

`ripr` targets the growth-rate optimal e-variable of safe testing [@GrunwaldDeHeideKoolen2024], approximating the reverse information projection (RIPr) of a fixed alternative `Q` onto a composite null `H0` that takes the form of a union of convex parts.

The idea it rests on is simpler than the projection itself. For any non-negative, bounded random variable `X`, if `b` is any number at least `sup_{theta in Theta_0} E_theta[X]`, then `X / b` has null expectation at most one. Specifically, `E = X / b` is an *e-variable*.

**This means the projection need not be fit exactly:** any approximation can
be used, so long as the resulting likelihood ratio is corrected against a
*proven* bound to yield a valid e-variable. The quality of the approximation
determines only how much e-power survives the correction. That splits a
hard, non-convex optimisation problem into an easy one (approximate) and a
proof (bound), and this split is what `ripr` facilitates: fitting is handled
by a Frank--Wolfe [@Jaggi2013] and EM hybrid over mixing measures, and
certification by a branch-and-bound algorithm (currently implemented for
multinomial families via the Bernstein enclosure).

This vignette walks through both halves on a worked example, then shows what
happens when the fit is deliberately bad, and finally contrasts a proven
bound against an uncertified search.

## A first null: the plurality null

Take a three-category multinomial distribution (`n = 20`) and the plurality
null: candidate 1 does not win outright,
`H0 = union_j {theta : theta_1 <= theta_j}`.
This null is a union of convex parts (one region where `theta_1 <= theta_2`
and another where `theta_1 <= theta_3`, each part a simplex), so the plurality
null is a single `null_model` holding two `simplex_region` parts.

```{r problem}
K <- 3L
n <- 20L
family <- multinomial_family(n_trials = n, k = K)

# Defines {theta : theta_1 <= theta_j}
plurality_part <- function(j) {
  vertices <- diag(K)
  vertices[, 1L] <- replace(numeric(K), c(1L, j), 0.5)
  simplex_region(vertices = vertices)
}


plurality <- null_model(
  family,
  union(plurality_part(2), plurality_part(3))
)
```

We nominate a discrete mixture-multinomial distribution as the alternative.
Here the mixing measure is a single dirac at `q = (0.40, 0.35, 0.25)`,
yielding a simple `Multinomial(20, q)` alternative.

```{r alternative}
Q <- family(c(0.40, 0.35, 0.25))
```

```{r plot-plurality, echo = FALSE, fig.cap = "Plurality null, with the support of the alternative mixing distribution W1."}
W1 <- Q@mixing
plot_null(
  plurality,
  points = atoms(W1),
  weights = weights(W1) / 4,
  bg = "white"
)
```

The two parts overlap, and the alternative sits in the cap above them,
where candidate 1 leads.

## Fitting the projection

The fitting procedure is a sequence of steps: `fw_step()` for Frank--Wolfe
updates and `em_step()` for EM (also `lb_step()` for Li--Barron
[@LiBarron1999], though it is not generally recommended -- see
`?oracles`). There is no fixed pipeline; which algorithm runs is decided by
which verbs are called, and in what order. Here we run 40 steps of
Frank--Wolfe, then use `ripr_finish()` to solve the weights and drop any
low-mass atoms.

```{r fit}
set.seed(1)
state <- ripr_init(Q, plurality)
state <- fw_step(
  state,
  times = 40L,
  record_gap = TRUE,
  until = gap_below(1e-10)
)
fit <- ripr_finish(state, reoptimise = TRUE, identify = TRUE, record_gap = TRUE)

c(kl = fit$kl, gap = fit$gap_final, atoms = n_atoms(fit$W0))
```

```{r plot-plurality-fit, echo = FALSE, fig.cap = "Plurality null, with the support of W1 (alternative) and W0 (null, fitted)."}
plot_null(
  plurality,
  points = list(atoms(fit$W0), atoms(W1)),
  weights = list(weights(fit$W0), weights(W1) / 4)
)
```

The fit attains a KL divergence (from `Q`) of approximately
`r round(fit$kl, 4)`. The `gap` quantity is a Frank--Wolfe duality gap:
at its true value it bounds the expectation of `Q / P*` over the null by
`1 + gap`. The `gap` recorded *during* a fit (`state@trace$gap`, or
`fit$gap_fit`) is only an estimate, though -- it is not itself an upper
bound, so it cannot be used directly to certify an e-value. `fit$gap_final`,
requested here via `record_gap = TRUE`, is a fresh sweep over the *returned*
mixture, and is what we compare against the true certified bound below.

## From a fit to an e-variable

With a fitted projection in hand, we can construct a `random_variable` that
computes the mixture likelihood ratio `Q / P*` -- our candidate e-variable:

```{r ratio}
X <- likelihood(Q, label = "Q") /
  likelihood(fit$P_star, label = "P*")
X
```

`random_variable`s are callable: they map elements of the sample space to
their realisation.

```{r rv_outcome}
outcomes <- rbind(
  c(10L, 10L, 0L),
  c(8L, 7L, 5L)
)
X(outcomes)
```

`certify()` proves an upper bound on the null expectation of `X`. It returns
the proven upper bound alongside the largest value its internal search
actually attained; the two bracket the true supremum, which is generally
unknown.

```{r certify}
cert <- certify(X, plurality, tol = 1e-9)
c(
  upper = cert$sup_ub,
  attained = cert$sup_lb,
  width = cert$sup_ub - cert$sup_lb
)
```

The bound lands just above one: it is exactly `1 + gap`, where `gap` is the
*true* Frank--Wolfe duality gap that the fit stopped on.

```{r identity}
c(bound_minus_one = cert$sup_ub - 1, fitted_gap = fit$gap_final)
```

The rescaled random variable `E = X / cert$sup_ub` is then a genuine
e-variable for `plurality`:

```{r evariable}
E <- X / cert$sup_ub
print(E)
print(E(outcomes))
```

## The correction doesn't need a good fit

The whole point of splitting fitting from certification is that validity
does not depend on the quality of the fit. To demonstrate, replace the
projection with something that is not the RIPr at all -- a uniform point
mass -- and certify the resulting ratio:

```{r bad-fit}
bad <- family(rep(1 / K, K))
X_bad <- likelihood(Q, label = "Q") /
  likelihood(bad, label = "P0")
cert_bad <- certify(X_bad, plurality, tol = 1e-9)

c(good = cert$sup_ub, bad = cert_bad$sup_ub)
```

Both rescale to valid e-variables. What differentiates them is e-power --
the log-growth, or expected log e-value under the alternative, that
compounds over a sequential procedure:

```{r growth}
growth <- function(E, Q) {
  outcomes <- enumerate_space(E@sample_space)
  e_values <- E(outcomes)
  Q_expectation <- log_density(Q, outcomes) |>
    exp() %*%
    log(e_values)
  c(Q_expectation) # return as scalar
}

E_gr <- growth(E, Q)

c(good = E_gr, bad = growth(X_bad / cert_bad$sup_ub, Q))
```

A negative growth rate means evidence is lost on average even when the
alternative is true -- useless, but not invalid. **Validity is free from the
certification and correction procedure; fitting a good approximation is what
buys e-power.**

When `P*` is the true RIPr of `Q` onto the null, the attained KL divergence
`KL(Q || P*)` is precisely the growth rate of the resulting e-variable. For
any fitted `P` the KL divergence sits above that optimum, so `fit$kl` upper
bounds the optimal growth rate. The true optimum therefore lies somewhere
between the growth this fit achieved and `fit$kl`:

```{r growth-kl-gap}
c(
  lower = E_gr,
  upper = fit$kl,
  width = fit$kl - E_gr
)
```

## A second worked example: composing step verbs

Because `fw_step()`, `lb_step()`, `em_step()` and `weight_step()` are
independent verbs rather than a fixed pipeline, they compose freely. Here we
demonstrate a null where candidate 1, 2 and 3 all fail to reach a majority
(`H0: theta_1, theta_2, theta_3 <= 1/2`, the medial triangle), fit with a
loop that interleaves Frank--Wolfe and EM steps, and nominate an alternative
with one dirac per section, equally weighted.

```{r medial}
medial <- null_model(
  family,
  simplex_region(
    vertices = cbind(c(0.5, 0.5, 0), c(0.5, 0, 0.5), c(0, 0.5, 0.5))
  )
)

W_maj <- finite_dist(
  components = cbind(
    c(0.6, 0.2, 0.2),
    c(0.2, 0.6, 0.2),
    c(0.2, 0.2, 0.6)
  ),
  weights = c(1 / 3, 1 / 3, 1 / 3)
)
Q_maj <- family(W_maj)
```

```{r plot-medial-null, echo = FALSE, fig.cap = "Medial-triangle null, with the support of W_maj."}
plot_null(
  medial,
  points = atoms(W_maj),
  weights = weights(W_maj) / 4,
  bg = "white"
)
```

```{r medial-alternative}
hybrid_step <- function(state, times) {
  for (iter in seq_len(times)) {
    state <- state |> fw_step() |> em_step()
  }
  state
}

set.seed(2)
fit_medial <- ripr_init(Q_maj, medial) |>
  hybrid_step(times = 40L) |>
  ripr_finish(reoptimise = TRUE, identify = TRUE, record_gap = TRUE)
X_medial <- likelihood(Q_maj, label = "Q") /
  likelihood(fit_medial$P_star, label = "P*")

c(kl = fit_medial$kl, upper = certify(X_medial, medial, tol = 1e-9)$sup_ub)
```

```{r plot-medial-fit, echo = FALSE, fig.cap = "Medial-triangle null, with the support of W_maj and W0 (fitted)."}
plot_null(
  medial,
  points = list(atoms(fit_medial$W0), atoms(W_maj)),
  weights = list(weights(fit_medial$W0), weights(W_maj) / 4)
)
```

## Searching vs. proving

`sup_lb()` hunts for the largest null expectation by multi-start gradient
ascent. It is much cheaper than proving a bound, and it works on families
and geometries that have no implemented bounding procedure -- but it is
**not** a certificate. It reports a *lower* bound on the supremum: something
larger may sit somewhere gradient ascent could not reach.

```{r sup-lb}
c(searched = sup_lb(X, plurality)$sup_lb, certified = cert$sup_ub)
```

Where no bounding method has been implemented for a given combination of
family and region geometry, `certify()` refuses outright, rather than
silently returning something unproven:

```{r refuse, error = TRUE}
gaussian <- gaussian_family(dim = 2L)
null <- null_model(
  gaussian,
  halfspace_region(normal = c(1, -1), offset = 0)
)
Y <- likelihood(gaussian(c(0, 0)))
certify(Y, null)
```

`sup_lb()` still works here, since it needs no bounding method -- only a
search.

## Where to go next

The function reference lists every exported function and S7 class, with a
runnable example on each. A few starting points if you're extending rather
than just using the package:

- `?parametric_family` describes the interface a new observation model must
  implement: `compile_loglik()`, `score()` and `kernel_draw()`. `?sample_space`
  and `?convex_region` describe the two sets it maps between, and
  `?distribution` the laws `fam(theta)` returns.
- `vignette("regions")` tours the region interface and the set algebra
  (`union()`, `intersect()`, `setdiff()`, `disjoin()`, ...) used to build a
  null out of convex pieces. `?convex_region` and its generics (`chart()`,
  `contains()`, `project()`) describe the interface a new convex geometry must
  implement; `?simplex_region` and `?halfspace_region` are the two built-in
  examples covered there.
- `?quadrature`, `?exact_engine`, `?mc_engine` and `?gh_engine` describe how
  expectations under the alternative are approximated, and how to add a new
  rule.
- `?oracles` explains the Frank--Wolfe and Li--Barron oracles that
  `fw_step()` and `lb_step()` are built on, and when to prefer one over the
  other.

## References
